Reanalysis Step 5: Main Effects Outcome Analysis

### 2025-08-31 reanalysis-5-outcome_analysis.qmd
refactor to emmeans

```{r marginal-effects}
# Check if AMEs are cached
ame_cache_file <- file.path(reanalysis_data_dir, "ame_results_main_effects.rds")
force_recalc_ame <- TRUE  # Set to TRUE to force recalculation

if (file.exists(ame_cache_file) && !force_recalc_ame) {
  # Load cached AMEs
  ame_cache <- readRDS(ame_cache_file)
  ame_list <- ame_cache$ame_list
  ame_loaded_from_cache <- TRUE
  ame_cache_location <- ame_cache_file
} else {
  # Calculate AME for each imputed dataset then pool
  ame_loaded_from_cache <- FALSE
  ame_list <- list()

  # Initialize progress tracking
  progress <- init_progress(n_imputations, "Calculating AMEs")

@@

    imp_data <- imputed_datasets[[i]]
    imp_data$weights <- twang_weights[[i]]

    # Calculate AME with clustering and weights
    # marginaleffects handles survey designs well
    # the model already handled the clustering
    ame <- marginaleffects::avg_comparisons(
      model,
      variables = "ever_lapse_binary",
      type = "response",   # average difference in predicted probabilities
      newdata = imp_data,
      wts = "weights"
    )

    ame_list[[i]] <- ame
  }

  # Save to cache
  ame_cache <- list(
    ame_list = ame_list,
    n_imputations = n_imputations,
    timestamp = Sys.time()
  )
  saveRDS(ame_cache, ame_cache_file)
  ame_cache_location <- ame_cache_file
}

# Extract estimates and SEs for pooling
# avg_comparisons returns a data frame, so we need to extract the first row
ame_estimates <- sapply(ame_list, function(x) {
  if (is.data.frame(x)) {
    x$estimate[1]
  } else {
    x$estimate
  }
})

ame_ses <- sapply(ame_list, function(x) {
  if (is.data.frame(x)) {
    x$std.error[1]
  } else {
    x$std.error
  }
})

# Apply Rubin's rules for pooling
pooled_ame_est <- mean(ame_estimates)
within_var <- mean(ame_ses^2)
between_var <- var(ame_estimates)
total_var <- within_var + between_var * (1 + 1/n_imputations)
pooled_ame_se <- sqrt(total_var)

# Calculate FMI and df
fmi <- (between_var * (1 + 1/n_imputations)) / total_var
df_ame <- (n_imputations - 1) / fmi^2

# CI using t-distribution
t_crit <- qt(0.975, df_ame)
ci_lower <- pooled_ame_est - t_crit * pooled_ame_se
ci_upper <- pooled_ame_est + t_crit * pooled_ame_se

pooled_ame <- data.frame(
  Effect = "Average Marginal Effect of Lapse",
  AME = pooled_ame_est,
  SE = pooled_ame_se,
  CI_Lower = ci_lower,
  CI_Upper = ci_upper,
  FMI = fmi,
  Interpretation = paste0(
    "On average, lapsing from care increases the probability of VI/blindness by ",
    round(pooled_ame_est, 3)
  )
)

kable(pooled_ame |> select(-Interpretation), 
      caption = "Average marginal effect of lapsing",
      digits = 4)

# Display interpretation separately
ame_text <- pooled_ame$Interpretation
```

**AME cache status:** `r ifelse(ame_loaded_from_cache, paste("Loaded from", ame_cache_location), paste("Newly calculated and saved to", ame_cache_location))`

>>>>

```{r marginal-effects}
# Check if marginal means are cached
emmeans_cache_file <- file.path(reanalysis_data_dir, "emmeans_results_main_effects.rds")
force_recalc_emmeans <- TRUE  # Set to TRUE to force recalculation

if (file.exists(emmeans_cache_file) && !force_recalc_emmeans) {
  # Load cached results
  emmeans_cache <- readRDS(emmeans_cache_file)
  emmeans_list <- emmeans_cache$emmeans_list
  contrasts_list <- emmeans_cache$contrasts_list
  emmeans_loaded_from_cache <- TRUE
  emmeans_cache_location <- emmeans_cache_file
} else {
  # Calculate marginal means for each imputed dataset then pool
  emmeans_loaded_from_cache <- FALSE
  emmeans_list <- list()
  contrasts_list <- list()

  # Initialize progress tracking
  progress <- init_progress(n_imputations, "Calculating marginal means")

@@

    # Get the fitted model for this imputation
    model <- fitted_models[[i]]

    # Calculate emmeans on the response scale
    emm <- emmeans(model, 
                   ~ ever_lapse_binary,
                   type = "response",
                   weights = "proportional")

    # Get contrast (difference in probabilities)
    contrast <- pairs(emm, reverse = TRUE)  # Lapsed - Not Lapsed

    emmeans_list[[i]] <- as.data.frame(emm)
    contrasts_list[[i]] <- as.data.frame(contrast)
  }

  # Save to cache
  emmeans_cache <- list(
    emmeans_list = emmeans_list,
    contrasts_list = contrasts_list,
    n_imputations = n_imputations,
    timestamp = Sys.time()
  )
  saveRDS(emmeans_cache, emmeans_cache_file)
  emmeans_cache_location <- emmeans_cache_file
}

# Pool predicted probabilities for each group
# Extract probabilities for not lapsed (ever_lapse_binary = 0)
prob_not_lapsed <- sapply(emmeans_list, function(x) {
  x[x$ever_lapse_binary == 0, "prob"][1]
})
se_not_lapsed <- sapply(emmeans_list, function(x) {
  x[x$ever_lapse_binary == 0, "SE"][1]
})

# Extract probabilities for lapsed (ever_lapse_binary = 1)
prob_lapsed <- sapply(emmeans_list, function(x) {
  x[x$ever_lapse_binary == 1, "prob"][1]
})
se_lapsed <- sapply(emmeans_list, function(x) {
  x[x$ever_lapse_binary == 1, "SE"][1]
})

# Pool using Rubin's rules for each group
# Not lapsed
pooled_prob_not_lapsed <- mean(prob_not_lapsed)
within_var_not_lapsed <- mean(se_not_lapsed^2)
between_var_not_lapsed <- var(prob_not_lapsed)
total_var_not_lapsed <- within_var_not_lapsed + between_var_not_lapsed * (1 + 1/n_imputations)
pooled_se_not_lapsed <- sqrt(total_var_not_lapsed)

# Lapsed
pooled_prob_lapsed <- mean(prob_lapsed)
within_var_lapsed <- mean(se_lapsed^2)
between_var_lapsed <- var(prob_lapsed)
total_var_lapsed <- within_var_lapsed + between_var_lapsed * (1 + 1/n_imputations)
pooled_se_lapsed <- sqrt(total_var_lapsed)

# Calculate pooled risk difference
risk_diff <- pooled_prob_lapsed - pooled_prob_not_lapsed

# For risk difference SE, extract from contrasts
# Note: contrasts from emmeans pairs() gives odds ratios, not risk differences
# So we calculate risk difference SE using delta method or from the individual SEs
risk_diff_var <- total_var_lapsed + total_var_not_lapsed  # Conservative approximation
risk_diff_se <- sqrt(risk_diff_var)

# Calculate FMI and df for risk difference
between_var_diff <- var(prob_lapsed - prob_not_lapsed)
within_var_diff <- risk_diff_var - between_var_diff * (1 + 1/n_imputations)
fmi <- (between_var_diff * (1 + 1/n_imputations)) / risk_diff_var
df_diff <- (n_imputations - 1) / fmi^2

# CI using t-distribution
t_crit <- qt(0.975, df_diff)
ci_lower <- risk_diff - t_crit * risk_diff_se
ci_upper <- risk_diff + t_crit * risk_diff_se

# Create results data frame
pooled_emmeans_results <- data.frame(
  Group = c("Not Lapsed", "Lapsed", "Risk Difference"),
  Estimate = c(pooled_prob_not_lapsed, pooled_prob_lapsed, risk_diff),
  SE = c(pooled_se_not_lapsed, pooled_se_lapsed, risk_diff_se),
  CI_Lower = c(
    pooled_prob_not_lapsed - qt(0.975, df_diff) * pooled_se_not_lapsed,
    pooled_prob_lapsed - qt(0.975, df_diff) * pooled_se_lapsed,
    ci_lower
  ),
  CI_Upper = c(
    pooled_prob_not_lapsed + qt(0.975, df_diff) * pooled_se_not_lapsed,
    pooled_prob_lapsed + qt(0.975, df_diff) * pooled_se_lapsed,
    ci_upper
  )
)

# Display results
kable(pooled_emmeans_results,
      caption = "Pooled marginal means and risk difference from emmeans",
      digits = 4,
      col.names = c("Group", "Probability", "SE", "Lower CI", "Upper CI"))

# Store for later use
pooled_ame <- data.frame(
  Effect = "Average Marginal Effect of Lapse",
  AME = risk_diff,
  SE = risk_diff_se,
  CI_Lower = ci_lower,
  CI_Upper = ci_upper,
  FMI = fmi,
  Interpretation = paste0(
    "On average, lapsing from care increases the probability of VI/blindness by ",
    round(risk_diff, 3)
  )
)

# Display interpretation
ame_text <- pooled_ame$Interpretation
```

**Marginal means cache status:** `r ifelse(emmeans_loaded_from_cache, paste("Loaded from", emmeans_cache_location), paste("Newly calculated and saved to", emmeans_cache_location))`
